Program Transformations for miniKanren

### 2020-07-07 papers/aplas/text/intro.tex
Abstract
Edits

Control issues in partial deduction of logic programming language \pro have been studied before~\cite{leuschel2002logic}.
Unlike Prolog, where atoms in the rhs of a clause cannot be arbitrarily reordered without changing the semantics of a program, in miniKanren the subgoals of conjunction/disjunction can be freely switched.
This opens a yet another possibility for optimization, not taken into account by approaches initially developed in the context of conventional logic programming.
% The ideas described there make use of the deterministic search strategy of \pro.
% However in \mk it is allowed to reorder some relation calls within the goal for better performance.
% While sometimes conjunctive partial deduction gives great performance boost, sometimes it does not behave as well as it could have.

In this paper, we study issues which conjunctive partial deduction faces being applied for \mk.
We also describe a novel approach to partial deduction for relational programming, \emph{conservative partial deduction}.
We implemented this approach and compared it with the existing specialization system (\ecce) for several programs.
We report here the result of the comparison and discuss why some \mk programs run slower after specialization.

>>>>

Control issues in partial deduction of logic programming language \pro have been studied before~\cite{leuschel2002logic}.
Unlike Prolog, where atoms in the rhs of a clause cannot be arbitrarily reordered without changing the semantics of a program, in \mk the subgoals of conjunction/disjunction can be freely switched.
This opens a yet another possibility for optimization, not taken into account by approaches initially developed in the context of conventional logic programming.
% The ideas described there make use of the deterministic search strategy of \pro.
% However in \mk it is allowed to reorder some relation calls within the goal for better performance.
% While sometimes conjunctive partial deduction gives great performance boost, sometimes it does not behave as well as it could have.

In this paper, we study issues which conjunctive partial deduction faces being applied for \mk.
We also describe a novel approach to partial deduction for relational programming, \emph{conservative partial deduction}.
We implemented this approach and compared it with the existing specialization system (\ecce) for several programs.
We report here the results of the comparison and discuss why some \mk programs run slower after specialization.

### 2020-07-07 papers/aplas/text/related.tex
Abstract
Edits

The heart of supercompilation-based techniques is \emph{driving}~---~a symbolic execution of a program through all possible execution paths.
The result of driving is so-called \emph{process tree} where nodes correspond to \emph{configurations} which represent computation state.
For example, in the case of pure functional programming languages, the computational state might be a term.
Each path in the tree corresponds to some concrete program execution.
The two main sources for supercompilation optimizations are aggressive information propagation about variables' values, equalities and disequalities, and precomputing of all deterministic semantic evaluation steps.
The latter process, also known as \emph{deforestation}~\cite{deforestation}, means  combining of consecutive process tree nodes with no branching.
When the tree is constructed, the resulting, or \emph{residual}, program can be extracted from the process tree by the process called \emph{residualization}.
Of course, process tree can contain infinite branches.
\emph{Whistles} --- heuristics to identify possibly infinite branches --- are used to ensure supercompilation terminates.
If a whistle signalls during the construction of some branch, then something should be done to ensure termination.
The most common approaches include stop driving the infinite branch completely (no specialization is done in this case and the source code is blindly copied into the residual program) or fold the process tree to a \emph{process graph}.
The main instrument to perform such a folding is \emph{generalization}.
Generalization, abstracting away some computed data about the current term, makes folding possible.
%  i.e. abstracting away some computed data about the current term which makes folding possible.
One source of infinite branches is consecutive recursive calls to the same function with an accumulating parameter: by unfolding such a call further one can only increase the term size which leads to nontermination.
The accumulating parameter can be removed by replacing the call with its generalization.
There are several ways to ensure process correctness and termination, most-specific generalization
(anti-unification) and using \emph{homeomorphic embedding}~\cite{Higman52,Kruskal60} as a
whistle being the most common.
%the most common being \emph{homeomorphic embedding}~\cite{Higman52,Kruskal60} used as a whistle and most-specific generalization of terms.
% When two dangerously similar nodes are encountered,
% For example, two consecutive recursive calls of a function with accumulating parameter, and the first is not an instance of the second, then one may construct a new, generalized, node such that both of original nodes are instances of the last one, then one of the initial nodes is replaced with the generalized one.

While supercompilation generally improves the behaviour of input programs and distillation can even provide superlinear speedup, there are no ways to predict the effect of specialization on a given program in the general.
What is worse, the efficiency of residual program from the target language evaluator point of view is rarely considered in the literature.
The main optimization source is computing in advance all possible intermediate and statically-known semantics steps at program transformation-time.
Other criteria, like the size of the generated program or possible optimizations and execution cost of different language constructions by the target language evaluator, are usually out of consideration~\cite{jonesbook}.
It is known that supercompilation may adversely affect GHC optimizations yielding standalone compilation more powerful~\cite{SCBE,TCES} and cause code explosion~\cite{SCHC}.
Moreover, it may be hard to predict the real speedup of any given program on concrete examples even disregarding the problems above because of the complexity of the transformation algorithm.
The worst-case for partial evaluation is when all static variables are used in a dynamic context, and there is some advice on how to implement a partial evaluator as well as a target program so that specialization indeed improves its performance~\cite{jonesbook,bulyonkov84}.
There is a lack of research in determining the classes of programs which transformers would definitely speed up.

@@

An empitical study shown that the most well-behaved strategy of local control in CPD for \pro is \emph{deterministic unfolding}~\cite{leuschel1997advanced}.
An atom is unfolded only if precisely one suitable clause head exists for it with the one exception: it is allowed to unfold an atom non-deterministically once for one local control tree.
This means that if a non-deterministic atom is the leftmost within a conjunction, it is most likely to be unfolded and to introduce many new relation calls within the conjunction.
We believe this is the core problem with CPD which limits its power when applied to \mk.
The strategy of unfolding atoms from left to right is reasonable in the context of \pro because it mimics the way programs in \pro execute.
But it often leads to larger global control trees and, as a result, bigger, less efficient programs.
The evaluation result of a \mk program does not depend on the order of atoms (relation calls) within a conjunction, thus we believe a better result can be achieved by selecting a relation call which can restrict the number of branches in the tree.
We describe our approach which implements this idea in the next section.

>>>>

The heart of supercompilation-based techniques is \emph{driving}~---~a symbolic execution of a program through all possible execution paths.
The result of driving is so-called \emph{process tree} where nodes correspond to \emph{configurations} which represent computation state.
For example, in the case of pure functional programming languages, the computational state might be a term.
Each path in the tree corresponds to some concrete program execution.
The two main sources for supercompilation optimizations are aggressive information propagation about variables' values, equalities and disequalities, and precomputing of all deterministic semantic evaluation steps.
The latter process, also known as \emph{deforestation}~\cite{deforestation}, means  combining of consecutive process tree nodes with no branching.
When the tree is constructed, the resulting, or \emph{residual}, program can be extracted from the process tree by the process called \emph{residualization}.
Of course, process tree can contain infinite branches.
\emph{Whistles} --- heuristics to identify possibly infinite branches --- are used to ensure supercompilation terminates.
If a whistle signalls during the construction of some branch, then something should be done to ensure termination.
The most common approaches are either to stop driving the infinite branch completely (no specialization is done in this case and the source code is blindly copied into the residual program) or to fold the process tree to a \emph{process graph}.
The main instrument to perform such a folding is \emph{generalization}.
Generalization, abstracting away some computed data about the current term, makes folding possible.
%  i.e. abstracting away some computed data about the current term which makes folding possible.
One source of infinite branches is consecutive recursive calls to the same function with an accumulating parameter: by unfolding such a call further one can only increase the term size which leads to nontermination.
The accumulating parameter can be removed by replacing the call with its generalization.
There are several ways to ensure process correctness and termination, most-specific generalization
(anti-unification) and using \emph{homeomorphic embedding}~\cite{Higman52,Kruskal60} as a
whistle being the most common.
%the most common being \emph{homeomorphic embedding}~\cite{Higman52,Kruskal60} used as a whistle and most-specific generalization of terms.
% When two dangerously similar nodes are encountered,
% For example, two consecutive recursive calls of a function with accumulating parameter, and the first is not an instance of the second, then one may construct a new, generalized, node such that both of original nodes are instances of the last one, then one of the initial nodes is replaced with the generalized one.

While supercompilation generally improves the behaviour of input programs and distillation can even provide superlinear speedup, there are no ways to predict the effect of specialization on a given program in general.
What is worse, the efficiency of residual program from the target language evaluator point of view is rarely considered in the literature.
The main optimization source is computing in advance all possible intermediate and statically-known semantics steps at program transformation-time.
Other criteria, like the size of the generated program or possible optimizations and execution cost of different language constructions by the target language evaluator, are usually out of consideration~\cite{jonesbook}.
It is known that supercompilation may adversely affect GHC optimizations yielding standalone compilation more powerful~\cite{SCBE,TCES} and cause code explosion~\cite{SCHC}.
Moreover, it may be hard to predict the real speedup of any given program on concrete examples even disregarding the problems above because of the complexity of the transformation algorithm.
The worst-case for partial evaluation is when all static variables are used in a dynamic context, and there is some advice on how to implement a partial evaluator as well as a target program so that specialization indeed improves its performance~\cite{jonesbook,bulyonkov84}.
There is a lack of research in determining the classes of programs which transformers would definitely speed~up.

@@

An empitical study shown that the most well-behaved strategy of local control in CPD for \pro is \emph{deterministic unfolding}~\cite{leuschel1997advanced}.
An atom is unfolded only if precisely one suitable clause head exists for it with the one exception: it is allowed to unfold an atom non-deterministically once for one local control tree.
This means that if a non-deterministic atom is the leftmost within a conjunction, it is most likely to be unfolded and to introduce many new relation calls within the conjunction.
We believe this is the core problem of CPD which limits its power when applied to \mk.
The strategy of unfolding atoms from left to right is reasonable in the context of \pro because it mimics the way programs in \pro execute.
But it often leads to larger global control trees and, as a result, bigger, less efficient programs.
The evaluation result of a \mk program does not depend on the order of atoms (relation calls) within a conjunction, thus we believe a better result can be achieved by selecting a relation call which can restrict the number of branches in the tree.
We describe our approach which implements this idea in the next section.
